---
title: "CEB preliminary"
output: word_document
---

```{r setup, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
opts_knit$set(root.dir = '../')
#global packages and plot labels
library(visreg)
library(multcomp)
library(car)
library(lme4)
library(MuMIn)

```

Below are the preliminary analyses from the CEB data.  I have included lots of the code so you can see how what the model setup looks like and what steps I went through. I basically just analysed CEB thickness by life form and site. Then plotted against every variables. I did not find anything particularly meaningful.

```{r datasets, warning=FALSE, message=FALSE}
source("master_scripts/plot_objects.R")

ceb <- read.csv("raw_data/ceb_thickness.csv")
niche <- read.csv("raw_data/species_niches.csv")

ceb2 <- merge(ceb, niche)
  ceb2$niche2 <- gsub("climber", "hemi-epiphyte", ceb2$niche)
  ceb2$niche2 <- as.factor(ceb2$niche2)
  #reorder from ground to canopy 
  ceb2$niche2<-factor(ceb2$niche2, 
               levels=c("terrestrial", "hemi-epiphyte", "epiphyte"))
  ceb2$curve_id <- with(ceb2, paste(genusspecies, plant_no, sep="-"))
#drop asplenium #6 with 0.0
cebnozero <- ceb2[-16,] 

#data for plotting
#larger trait dataset
ferndat <- read.csv("calculated_data/ferns_traits_complete.csv")

ceb_regress <- merge(ceb2, ferndat, all=TRUE)
ceb_regress2 <- ceb_regress[complete.cases(ceb_regress$ceb_thickness_um),]

#k max data
kmax <- read.csv("calculated_data/lascruces_kmax.csv")
kmax2 <- merge(kmax, ceb2)

#p50
p50 <- read.csv("calculated_data/species_p50.csv")
names(p50)[1] <- "curve_id"
p50dat <- p50[p50$x==50,]
p50_ceb <- merge(ceb2[,c(1,4:5,7,9:10)], p50dat)

```

## CEB by life form

```{r ceb, warning=FALSE, message=FALSE}
boxplot(ceb_thickness_um ~ niche2, data=ceb2) 
#data check
hist(sqrt(ceb2$ceb_thickness_um), main="Square-root transformed for normal distribution")

#test interaction with site and if the indiviudal with 0.0 ceb matters
ceb_mod <- lmer(sqrt(ceb_thickness_um) ~ niche2 * site + (1|species), 
                   data=ceb2)
ceb_mod2 <- lmer(sqrt(ceb_thickness_um) ~ niche2 + site + (1|species), 
                    data=ceb2)
#without asp-6
ceb_mod3 <- lmer(sqrt(ceb_thickness_um) ~ niche2 * site + (1|species), 
                 data=cebnozero)

#model diagnostics
qqPlot(residuals(ceb_mod)) #good
plot(ceb_mod) ##transformation fixes skewness

#model summary
#Anova(ceb_mod3, type="3") #no interactions
#anova(ceb_mod, ceb_mod2) #models similar (with and without interaction)
print(AIC(ceb_mod, ceb_mod2)) #keep models with interaction

#effect of life form (niche2) present
print(Anova(ceb_mod, type="3"))
print(r.squaredGLMM(ceb_mod)) #diff between two r2 values is the variation of CEB in species, so here quite a lot

#Tukeys test for model with and without asplenium-6
tukey_ceb <- glht(ceb_mod, linfct = mcp(niche2 = "Tukey"))
print(cld(tukey_ceb))

tukey_ceb_noasp <- glht(ceb_mod3, linfct = mcp(niche2 = "Tukey"))
print(cld(tukey_ceb_noasp))

#hemi-epiphyte is different from epiphyte, probably not that exciting. See box plot above. Aspelnium-6 doesnt change results
```

## Bivariate plots with CEB

I just assumed we wanted to treat CEB thickness as the independent variable.

### CEB vs pressure volume curve parameters
```{r tlp}
par(mar=c(4,4,1,1))
plot(waterpot_tlp ~ ceb_thickness_um,pch=16, col=trtcols2[niche2],
     data=ceb_regress2)
plot(osmotic_potential ~ ceb_thickness_um,pch=16, col=trtcols2[niche2],
     data=ceb_regress2)
plot(capacitance_full ~ ceb_thickness_um,pch=16, col=trtcols2[niche2],
     data=ceb_regress2)
plot(elasticity ~ ceb_thickness_um,pch=16, col=trtcols2[niche2],
     data=ceb_regress2)
```

### CEB vs basic morphology
```{r morph,}
#basic plots morphology
plot(lamina_area_cm2 ~ ceb_thickness_um,pch=16, col=trtcols2[niche2],
     data=ceb_regress2)
plot(frond_length_cm ~ ceb_thickness_um,pch=16, col=trtcols2[niche2],
     data=ceb_regress2)
plot(stipe_length_cm ~ ceb_thickness_um,pch=16, col=trtcols2[niche2],
     data=ceb_regress2)
plot(elasticity ~ ceb_thickness_um,pch=16, col=trtcols2[niche2],
     data=ceb_regress2)
plot(sla_cm2g ~ ceb_thickness_um,pch=16, col=trtcols2[niche2],
     data=ceb_regress2)
```

### CEB vs internal anatomy
```{r anat}
plot(xylem_area_mm2 ~ ceb_thickness_um,pch=16, col=trtcols2[niche2],
     data=ceb_regress2)
plot(stomatal_size ~ ceb_thickness_um,pch=16, col=trtcols2[niche2],
     data=ceb_regress2)
plot(sd_mm2 ~ ceb_thickness_um,pch=16, col=trtcols2[niche2],
     data=ceb_regress2)
```

### CEB vs leaf chemistry
```{r chem}
#basic plots chemistry
plot(d13C ~ ceb_thickness_um,pch=16, col=trtcols2[niche2],
     data=ceb_regress2)
plot(n_perc ~ ceb_thickness_um,pch=16, col=trtcols2[niche2],
     data=ceb_regress2)
```

### CEB vs V-Curve parameters
```{r}
plot(K ~ ceb_thickness_um,pch=16, col=trtcols2[niche2],
     data=kmax2)

plot(Px ~ ceb_thickness_um,pch=16, col=trtcols2[niche2],
     data=p50_ceb, ylab="p50")
```

## Bivariate stats
I combed through all of these and this is the only significant relationship I found with the full model

```{r stats}
elas_ceb_mod <- lmer(elasticity ~ ceb_thickness_um  * niche2 +(1|species), data=ceb_regress2)
print(Anova(elas_ceb_mod, type=3))
#elasticity varies by CEB, but 2 data points drive it
visreg(elas_ceb_mod, "ceb_thickness_um", by="niche2")

#pairwise contrasts show how hemi is different (but the plot above makes this spurious)
library(emmeans)
emmeans(elas_ceb_mod, pairwise ~ ceb_thickness_um : niche2)

```

