---
title: "Tropical Fern Leaf Water Relations"
output: html_document
---

```{r, results="asis", echo=FALSE, warning=FALSE}
library(knitr)
opts_knit$set(root.dir = '../')
```

# Background

Below is the compilation of the working data sets we plan to use for the manuscript. Eventually the paper will inclide findings from the following traits (traits in **bold** are analyzed):

* PV Curves
* Stomata: size and **density**
* Vein density
* **SLA**
* **Foliar Nitrogen**
* **Isotopes**
* **Morphometrics**
* Phylogeny
* **Chlorophyll**

# Data Analysis

I have set up a linear mixed effect model design for each trait analysis. The general model equation is:

\

**trait ~ niche2 x site + (1|species)**

\

Niche2 re-codes our one climbing species as a terrestrial species. In this setup, species is used a a random effect, since individual species effects are not one of our central questions. This model returns 2 R^2^ values, one for the whole model and one for the fixed effect only. This allows us to evaluate the amount of variation in a trait attributed solely to species differences.

\

First, I check for the presence of an interaction. Most of the time there is not one. I also re-run the model with only niche2 and site as main effects, and then evaluate the models against each other (AIC scores). In most cases, I drop the interaction term and use the simplest model. Depending on our question, this setup can change. For example, if we have no interest in site effects, I can embed species within site and treat them both as random effects. Currently, I am using a type 3 ANOVA because we have a unbalanced design (hemi-epiphytes)

# Data visualizations

Below are a series of figures visualizing traits between terrestrial, epiphytic and hemi-epiphytic tropical ferns. I have overlayed the raw data on top of box plots for each treatment. The width or each box plots is proportional to the square-root of the number of observations in each group. If there are significant main effects ($\alpha$ level of 0.05) I have included significance letters from a post-hoc Tukey's test.

## Frond Morphology

```{r morph,  echo=FALSE, fig.align='center',fig.width = 6, fig.height = 6, warning=FALSE}
source("master_scripts/morphology_plots.R")
```
No differences in lamina area, however, this depends on which model I choose (with or without niche x site interaction).  There is a massive amount of variation tied up in species-specific differences. Stipes in terrestrial species are 56% longer than epiphytic or hemi-epiphytic species. Total frond length is 30% longer in terrestrial than epiphytic species, with hemi-epiphytic species as intermediates.
\

```{r bivar,  echo=FALSE, fig.align='center',fig.width = 6, fig.height = 6}
source("master_scripts/frondlength_laminaarea_plot.R")
```
Not surprising, but frond area - length relationships are strong within across all fern species.

## Stomatal Traits

```{r sd,  echo=FALSE, fig.align='center',fig.width = 6, fig.height = 6,warning=FALSE}
source("master_scripts/stomatal_density_plot.R")
```
Epiphytic and hemi-epiphytic species have 54% fewer stomata per given area than terrestrial species!

\

```{r ss,  echo=FALSE, fig.align='center',fig.width = 6, fig.height = 10, warning=FALSE}
source("master_scripts/stomata_size_plot.R")
```
Our undergraduate researcher as pushed through Juliette's beautiful pictures from Las Cruces. He has just now started Alex and Court's pictures from La Selva (pray for him). I have not analyzed these yet. 


## Vein Density

```{r vd,  echo=FALSE, fig.align='center',fig.width = 6, fig.height = 6}
source("master_scripts/vein_density_plots.R")
```


## Foliar Chemistry 

```{r nitro,  echo=FALSE, fig.align='center',fig.width = 6, fig.height = 6}
source("master_scripts/leaf_chemistry_plots.R")
```
Foliar nitrogen is 29% lower in epiphytes than terrestrial species, with hemi-epiphytes as intermediates.

\

```{r 13c,  echo=FALSE, fig.align='center',fig.width = 10, fig.height = 6}
source("master_scripts/13C_plots.R")
```
Mean foliar $\delta$^13^C for epiphytes is -34.4&permil; compared with -32.5&permil; for terrestrial + hemi-epiphytes. These is also a significant site effect. Is this relevant? 

\

```{r 15n,  echo=FALSE, fig.align='center',fig.width = 6, fig.height = 6}
source("master_scripts/15N_plots.R")
```
There is an interaction between site and niche that is likely not biologically relevant; some treatments for niche are lower at Las Cruces (but not all).  

# Chlorophyll Content

```{r chl,  echo=FALSE, fig.align='center',fig.width = 6, fig.height = 6}
source("master_scripts/chlorophyll_plots.R")
```
I had to perform a little data mining to get this result. I dropped one epiphytic species with way to variable data (400-800 mg m^-2^) and a few outliers. Not sure how meaningful these differences are.

# Foliar Hydrauliic Traits

Below are visualizations for some of the hydraulic traits we derived from pressure volume curves. These traits are calculated from Sacks's excel macro. I am worried that I have messed up 2 calculations from the macro when I first ran La Selva curves. Since I need to check them thoroughly, I only show Las Cruces below. I will run stats once I am confident with La Selva data, but it looks like we may have some niche differences emerging.

```{r pv,  echo=FALSE, fig.align='center',fig.width = 6, fig.height = 6}
source("master_scripts/pv_plots.R")
```

