---
title: "Leaf water relations in epiphytic ferns are driven by avoidance rather than tolerance mechanims "
author: 
date: 
output:
  word_document:
    reference_docx: manuscriptstyle.docx
  pdf_document: default
  html_document:
    number_sections: yes
csl: new-phytologist.csl
bibliography: epi_leaf_zotero.bib
---

\
Courtney E. Campany^1^, Alex Baer^2^, Helen Holmlund^3^, Jarmila Pitterman^2^ and James E. Watkins Jr. ^1^

\
^1^ Department of Biology, Colgate University, Hamilton, NY, USA

\
^2^  Department of Ecology and Evolutionary Biology, University of California, Santa Cruz, California 95064, USA

\
Corresponding author: C.E. Campany, email: courtneycampany@gmail.com, 

\
Running Head: Terrestrial and epiphtic fern functional traits

```{r, results="asis", echo=FALSE, warning=FALSE}
library(knitr)
opts_knit$set(root.dir = '../')
```

```{r global data/sourcing, echo=FALSE, message=FALSE, warning=FALSE}
# Set Flag to include figures in text or not.
includeFigs <- TRUE
```

```{r figtabs, echo=FALSE}


```
# Abstract

\

Key Words:

# Introduction

Within forests, epiphytes grow at many levels. Some
are restricted to the dark understory, whereas others grow on
the exposed twigs of emergent tree (Watkeins 2012)

Kohn 2010).
These data suggest that terrestrial ferns have reduced water
use efficiency, which would be consistent with Brodribb and
McAdam’s hypothesis of inefficient stomatal control

Within the Polypod clade, there is a fundamental split, with one clade radiating into the canopy and a sister clade diversifying on the forest floor (fig. 2). This split takes place between the eupolypod I clade that is predominantly epiphytic and the eupolypod II that is more than 90% terrestrial (estimates derived from counts in Moran and Riba 1995 and Mickel and Smith 2004). Within the eupolypod II clade, were it not for Aspleniaceae, there would be essentially no epiphytes in this lineage. (watkins 2012).

# Methods

## Study Site and Species Selection

The sites used for this study included two Costa Rican wet tropical forest locations at La Selva Biological Research Station in Heredia (84⁰00’12W, 10⁰25’52N) and Las Cruce Research Station in San Vito (8° 47′ 7” N, 82° 57′ 32” W). The La Selva site is a low elevation (ca 50 m) tropical forest, with a moderate dry season. The Las Cruces site is a premontane tropical forest located at a higher elvation (ca 1200 m). Both sites receive approximately 4m of annual rainfall [@holdridge_life_1967,@gentry_four_1993].

\

A survey of morphological, stoichiometric, anatomical and leaf water relations parameters were conducted for six individuals from 39 fern species across three fundamentally distinct life froms (Table 1). Across both sites, 18 terrestrial, 15 epiphytic and 6 hemi-epiphytic species were collected and meausured.  In this study, terrestrial life froms were all collected from shaded closed canopy understories in the forest floor. Epiphyitc life forms were sampled from trunks or within tree canopies,depending on the species. Epiphytic species were collected from canopy trees using single-rope climbing techniques. Hemi-epiphytic species were all collected along lower sections of trees trunks (1-3 m).  Importantly, all sampled hemi-epiphtic species are known to have root connections to forest floor soils at some point in their life history. Individuals of species were collected across multiple populations but within similar microhabitat conditions. All sampled fern species were restricted to the eupolypod I and II clades. Vouchers for each species were deposited at the sites of collection at either the La Selva (LSCR) or Las Cruces (LCCR) herbariums. 

## Plant Material

Two complete fronds from sampled individuals were field collected in the early morning (6-7:00 am). Stipes were cut at the base of the rhizome and cut ends were wrapped in wet paper towels and transported to the lab in black plastic bags.  Stipes were re-cut under water adn rehydrated until time of hydaulic measurement (1-6 hours). Due to the difficulty in sampling some canopy species; whole epiphytic individuals were removed, maintained overnight in well-watered conditions in an abient air laboratory and sampled the following day. One frond from each individual was utlized for presssure volume curves, while the other was sampled for strucutral morphology, lamina stoichmetry and anatomical traits. 
## Leaf Morphometric traits

Stipe length (cm) and lamina length (cm) were calculated from one sampled frond per indiviudal. Total lamina area for each frond was measured with a Li-3100 leaf area meter (LiCor Biosciences, ). Leaf mass per unit area (LMA, g cm^-2^) was calculated using the tissue punch method. For each individual, ten lamina punches (5 mm ^2^) were dried to a constant mass and LMA was calculated as the total dry mass divided by the total area of all leaf punches. 

## Anatomical traits

Total **xylem vascular** area was determined for stipe segments from hand cross-sections. Each cross section was stained with toulune blue....for. 
Images under 10x were ...
Total vascular area was determined by measuring the area of all vascular bundles within a stipe cross section. Images were analyzed using Image J software (National Institutes of Health, Bethesda Maryland, USA). 

\

Stomatal density (SD) was measured by directly counting stomata on the abaxial leaf surface under 40x magnification. Three leaf punches (4 mm^2^ diameter) were sampled across random locations on different pinnae from each individual. The number of stomata in each field of fiew were counted in three random regions on each of three leaf punches. The stomatal density for each indivudal is presented as the mean SD across all 9 sampled regions

\

Indiviudal images of stomata were directly photographed under 40x magnification (AmCcope FMA050) across all three leaf punches per individual. Stomatal length and width of both guard cells were calculated for 9 stomata for each individual using Image J. Stomatal size (SS) was calculated as guard cell length multipled by the combined width of each guard cell pair, as in @franks_maximum_2009.

## Foliar chemistry

Subsamples of foliar tissue used for lamina area calculation were samples across muliple pinnae for each individual.  These subsamples were dired to a constant mass and ground using a Wig-L-Bug (Sigma-Aldrich Co. St. Louis, USA).

Nitrogen content and deltaC^13^  were measured using a 

 by comparison with certified standards.

## Pressure volume curves

We used a modified pressure–volume technique of Tyree and Hammel (1972) to estimate the following tissue-water relations: osmotic potential at full tissue hydration ( Ψ π , sat ), osmotic potential at the turgor loss point of tissues ( Ψ π , tlp ), bulk modulus of tissue elasticity ( ε ), and tissue-water capacitance ( C ). This method has been fully described by Davis and Mooney (1986). 

\

For each pressure-volume curve we sampled top most intact pinnae for each fully hydrated frond. We generated pressure–volume curves by taking sequential water potential measurements ( Ψ ) as fronds air dried first in closed plastic bages (0-2 hrs) and then on a bench top. The fresh mass was recored immediately before and after each Ψ determination. Following each PV curve, samples were dried to a constant mass to calculate relative water content (RWC).

\

Lawren Sack's  = something here
 To determine Ψπ, tlp , a plot of 1/ Ψ vs. RWC was generated, and the
point at which the graph linearized was taken as the Ψπ, tlp ( Davis and
Mooney, 1986 ).  Ψπ, sat was determined by extending the regression line to the point where x = 0 (RWC =100). ε was determined from the slope m of the line of Ψpvs. RWC and the RWC at the turgor loss point (RWC°), such that ε= m RWC ° ,
and C was determined as the slope of RWC vs. Ψ ( Table 2 ). 

## Statistical analysis

Tests of allometric relationships between log-transformed biomass components were implemented using standardized major axis regression in the 'smatr' package in R [@warton2012smatr]. All tests of statistical significance were conducted at an $\alpha$ level of 0.05.

#Results

## Frond Morphology

```{r morph,  echo=FALSE, fig.align='center',fig.width = 6, fig.height = 6, warning=FALSE}
source("master_scripts/morphology_plots.R")
```
No differences in lamina area, however, this depends on which model I choose (with or without niche x site interaction).  There is a massive amount of variation tied up in species-specific differences. Stipes in terrestrial species are 56% longer than epiphytic or hemi-epiphytic species. Total frond length is 30% longer in terrestrial than epiphytic species, with hemi-epiphytic species as intermediates.

\

```{r sla,  echo=FALSE, fig.align='center',fig.width = 6, fig.height = 6}
source("master_scripts/sla_plot.R")
```
Specific leaf area is 35% higher in terrestrials ferns compared to epiphytic ferns, with hemi-epiphytes as intermediates.

\

```{r bivar,  echo=FALSE, fig.align='center',fig.width = 6, fig.height = 6}
source("master_scripts/frondlength_laminaarea_plot.R")
```
Not surprising, but frond area-length relationships are strong across all fern species.

## Stomatal Traits

```{r sd,  echo=FALSE, fig.align='center',fig.width = 6, fig.height = 6,warning=FALSE}
source("master_scripts/stomatal_density_plot2.R")
```
Epiphytic and hemi-epiphytic species have 51% fewer stomata per given area than terrestrial species!

\

```{r ss,  echo=FALSE, fig.align='center',fig.width = 6, fig.height = 10, warning=FALSE}
source("master_scripts/stomata_size_plot.R")
```
Our undergraduate researcher as pushed through Juliette's beautiful pictures from Las Cruces. He has just now started Alex and Court's pictures from La Selva (pray for him). I have not analyzed these yet. Gurad cell width is the average of the two guard cells.

## Vein Density

```{r vd,  echo=FALSE, fig.align='center',fig.width = 6, fig.height = 6}
source("master_scripts/vein_density_plots.R")
```


## Foliar Chemistry 

```{r nitro,  echo=FALSE, fig.align='center',fig.width = 6, fig.height = 6}
source("master_scripts/leaf_chemistry_plots.R")
```
Foliar nitrogen is 29.8% lower in epiphytes compared to other groupgs

\


```{r nitro,  echo=FALSE, fig.align='center',fig.width = 6, fig.height = 6}
source("master_scripts/nitro_lma.R")
```
Foliar nitrogen and lma relationships differ between habitat groups

\

```{r 13c,  echo=FALSE, fig.align='center',fig.width = 10, fig.height = 6}
source("master_scripts/13C_plots.R")
```
Mean foliar $\delta$^13^C for epiphytes is -34.4&permil; compared with -32.5&permil; for terrestrial + hemi-epiphytes. These is also a significant site effect. Is this relevant? 

\

```{r 15n,  echo=FALSE, fig.align='center',fig.width = 6, fig.height = 6}
source("master_scripts/15N_plots.R")
```
There is only a site effect.  

## Chlorophyll Content

```{r chl,  echo=FALSE, fig.align='center',fig.width = 6, fig.height = 6}
source("master_scripts/chlorophyll_plots.R")
```
I had to perform a little data mining to get this result. I dropped one epiphytic species with way to variable data (400-800 mg m^-2^) and a few outliers. Not sure how meaningful these differences are.

## Foliar Hydraulic Traits

Below are visualizations for some of the hydraulic traits we derived from pressure volume curves. These traits are calculated from Sacks's excel macro. I am worried that I have messed up 2 calculations from the macro when I first ran La Selva curves. Since I need to check them thoroughly, I only show Las Cruces below. I will run stats once I am confident with La Selva data, but it looks like we may have some niche differences emerging.

```{r pv,  echo=FALSE, fig.align='center',fig.width = 6, fig.height = 6}
source("master_scripts/pv_plots.R")
```

```{r pv,  echo=FALSE, fig.align='center',fig.width = 6, fig.height = 6}
source("master_scripts/tlp_osmpot_plot.R")
```

```{r pv,  echo=FALSE, fig.align='center',fig.width = 6, fig.height = 6}
source("master_scripts/tlp_elasticity_plot.R")
```



# Discussion

(Brodribb 2017): A passive linkage between leaf water status and guard cell turgor observed in extant basal vascular plants appears to be sufficient to prevent xylem cavitation during diurnal changes in evaporative demand (Martins et al., 2015). However, without more sophisticated mechanisms to reduce guard cell turgor and produce complete stomatal closure, it has been hypothesized that passive closure does not provide a sufficiently tight stomatal seal capable of preventing ferns and lycophytes from rapidly reaching critical leaf water potentials when soil water is depleted during drought (McAdam and Brodribb, 2013). As a result, ferns and lycophytes in dry environments rely on either a high plant capacitance or low stomatal density (McAdam and Brodribb, 2013), desiccation tolerance (Hietz, 2010), and in some cases, rather cavitation-resistant xylem (Baer et al., 2016)