---
main_topsize: 0.2 #percent coverage of the poster
main_bottomsize: 0.1
#ESSENTIALS
title: '**Leaf water relations in epiphytic ferns are driven by avoidance rather than tolerance mechanisms**'
author:
  - name: Courtney Campany
    affil: 1
    main: true
    orcid: '0000-0002-1099-3857'
    twitter: Court_Campany
    email: ccampany@shepherd.edu
  - name: Jarmila Pittermann
    affil: 2
    main: true
  - name: Alex Baer
    affil: 2
  - name: Helen Holmlund
    affil: 3
  - name: Eddie Watkins
    affil: 4
affiliation:
  - num: 1
    address: Department of Biology, Shepherd University, Shepherdstown, WV, USA
  - num: 2
    address: Department of Ecology and Evolutionary Biology, University of California, Santa Cruz, CA, USA
  - num: 3
    address: Natural Science Division, Pepperdine University, Malibu, CA, USA
  - num: 4
    address: Department of Biology, Colgate University, Hamilton, NY, USA
main_findings:
  - "Make **better posters** with RMarkdown + **posterdown**."
logoleft_name: https&#58;//raw.githubusercontent.com/brentthorne/posterdown/master/images/betterhexlogo.png
logoright_name: https&#58;//raw.githubusercontent.com/brentthorne/posterdown/master/images/betterhexlogo.png
logocenter_name: https&#58;//raw.githubusercontent.com/brentthorne/posterdown/master/images/qr-code-black.png
output: 
  posterdown::posterdown_betterport:
    self_contained: false
    pandoc_args: --mathjax
    number_sections: false
bibliography: packages.bib
link-citations: true
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      tidy = FALSE,
                      message = FALSE,
                      fig.align = 'center',
                      out.width = "100%")
options(knitr.table.format = "html") 
```

# Introduction

Opportunistic diversification allowed ferns to radiate into epiphytic niches created by now angiosperm dominated landscapes. Compared to terrestrial habitats, ferns in epiphytic habitats should exhibit variation in functional traits to cope with water stress. *We tested whether the evolution of leaf traits related to drought tolerance promoted the radiation of ferns from moist, terrestrial habitats to more unpredictable canopy habitats.*

```{r, include=FALSE}
knitr::write_bib(c('posterdown', 'rmarkdown','pagedown'), 'packages.bib')
```

## Objectives

1. **Explore** epiphytic trait divergence in a phylogenetic context
2. **Examine** leaf responses to water stress vs functional traits
3. **Explain** opportunistic radiation of ferns in canopy habitats

# Methods

* **Surveyed 39 fern species in Costa Rican tropical forests**
* **Explore epiphytic functional trait divergence in a phylogenetic context**
* **_Now on to the results!_**

# Results

Here you may have some figures to show off, bellow I have made a scatterplot with the infamous Iris dataset and I can even reference to the figure automatically like this, `Figure \@ref(fig:irisfigure)`, Figure \@ref(fig:irisfigure).

```{r, irisfigure, fig.cap='Here is a caption for the figure. This can be added by using the "fig.cap" option in the r code chunk options, see this [link](https://yihui.name/knitr/options/#plots) from the legend himself, [Yihui Xie](https://twitter.com/xieyihui).', out.width="80%"}
par(mar=c(2,2,0,1))
plot(x = iris$Sepal.Length, y = iris$Sepal.Width,
     col = iris$Species, pch = 19, xlab = "Sepal Length",
     ylab = "Sepal Width")
```

Maybe you want to show off some of that fancy code you spent so much time on to make that figure, well you can do that too! Just use the `echo=TRUE` option in the r code chunk options, Figure \@ref(fig:myprettycode)!

```{r stomata, echo=TRUE, fig.cap='Boxplots, so hot right now!', fig.height=6, out.width="80%"}
source("plots/figure4.R")
```

How about a neat table of data? See, Table \@ref(tab:iristable):

```{r pvcurve, echo=TRUE, fig.cap='Epiphytic tropical ferns are not more drought tolerant than terrestrial or hemi-epiphytic tropical ferns', fig.height=6, out.width="80%"}
source("plots/figure5.R")
```

# References
